High geomagnetic field intensity recorded by anorthosite xenoliths requires a vigorous late Mesoproterozoic geodynamo

### 2021-12-18 PNAS_manuscript/PNAS_manuscript.tex
combine figure, code and text updates

% Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
\author[a,1]{Yiming Zhang}
\author[a,1]{Nicholas L. Swanson-Hysell} 
\author[a,b]{Margaret S. Avery}
\author[c]{Roger R. Fu}

@@

% Please add a significance statement to explain the relevance of your work
\significancestatement{Acquiring high-fidelity observational records of ancient magnetic field intensity from remanent magnetization of rocks is crucial for constraining the evolution history of Earth's core. However, such records are often difficult to recover due to the alterations associated with weathering after the formation of rocks and heating during experiments. In this study, we use plagioclase cumulates from the North American Midcontinent Rift that has experienced minimal post-formation alterations to obtain a present-day-like virtual dipole moment 1.1 billion years ago. The results indicate the existence of strong power sources for Earth's dynamo during the late Mesoproterozoic. }

% Please include corresponding author, author contribution and author declaration information
\authorcontributions{Author contributions: Y.Z. and N.L.S.-H. designed research; Y.Z., N.L.S.-H. and M.S.A studied and sampled the rock cores; Y.Z. and N.L.S.-H. conducted and analyzed rock magnetic experiments; Y.Z. performed petrography and electron microscopy; R.R.F. contributed to magnetic imaging data analyses; and Y.Z., N.L.S.-H., M.S.A, and R.R.F. wrote the paper.}

@@

\begin{abstract}
Obtaining estimates of Earth's magnetic field strength in deep time is complicated by non-ideal rock magnetic behavior in many igneous rocks. In this study, we target \textit{ca.} 1092 Ma anorthosite xenoliths that cooled and acquired their magnetization within the Beaver River diabase intrusions of the North American Midcontinent Rift. In contrast to the diabase which has non-ideal paleointensity behaviors leading to its failure in providing reliable paleointensity estimates, the anorthosite xenoliths often yield high-quality, single-slope paleointensity results that are consistent at specimen and site levels. An average value of $\sim$89 ZAm$^2$ for the virtual dipole moment from the anorthosites is higher than that of the dipole component of Earth's magnetic field today. Such a strong late Mesoproterozoic dipole recorded by the anorthosites require the existence of a vigorous geodynamo at the time. Together with previous paleointensity data from other Midcontinent Rift rocks, our new results suggest that a dynamo with strong power sources likely persisted through $\sim$25 Myr at \textit{ca.} 1.1 Ga. These observations are inconsistent with there being a progressive decay of Earth's dynamo strength through the Proterozoic Eon.
\end{abstract}

@@

% If your first paragraph (i.e. with the \dropcap) contains a list environment (quote, quotation, theorem, definition, enumerate, itemize...), the line after the list may have some extra indentation. If this is the case, add \parshape=0 to the end of the list environment.
\dropcap{E}arth's magnetic field is the result of convective flow of liquid iron-alloy in Earth's outer core. At present day, the geodynamo is collectively powered by heat flow across the core-mantle boundary (CMB), and from the crystallization of the solid inner core from the liquid outer core which provides latent heat and compositional buoyancy. However, while paleomagnetic studies have found that a dynamo field has existed since at least 3.4 billion years ago in the Archean \cite{Selkin2007a, Tarduno2014a, Brenner2020a}, Earth's inner core likely crystallized more recently. Estimates of the timing of the initial crystallization of Earth's inner core are interconnected with estimates for the core's thermal conductivity. Higher conductivity values imply faster cooling rates, which in turn imply that the threshold for the freezing of the inner core happened more recently \cite{Davies2015a}. While some estimates of these values are consistent with an inner core age $>$3 Ga \cite{Gubbins2004a, Konopkova2016a}, other estimates based on first principles calculations implied higher thermal conductivity values and an age for the inner core that is less than 1.3 Ga \cite{Pozzo2012a, Koker2012a, Gomi2013a, Zhang2020b}, with some suggesting even younger ages ($<$ 700 Ma; ref.s \citealp{Labrosse2015a, Ohta2016a}. With the estimates for the core's thermal conductivity continue to be debated, it is crucial to use observational records as an independent constraint on the thermal evolution of Earth's core and mantle.

@@

Paleomagnetic records from ancient rocks are one of the few types of observational data that have the potential to record the evolution of Earth’s core and can be used to test the predictions from numerical simulations. Interpretations of selected records of the current Precambrian absolute paleointensity database (compiled in the PINT database; \url{http://earth.liv.ac.uk/pint/}) suggest that an appreciable geomagnetic field existed in the Archean but reached its weakest in the Ediacaran when very low virtual dipole moment (VDM) estimates have been obtained \cite{Bono2019a, Shcherbakova2019a, Thallner2021b, Lloyd2021a}, followed by a recovery near the Ediacaran-Cambrian boundary \cite{Thallner2021a}. Ref. \citealp{Bono2019a} adopted a billion-year-scale view of such observations to suggest that these data are consistent with thermal evolution models that use high core thermal conductivity values, where the nucleation of the inner core did not begin to provide compositional convection power for driving the geodynamo until the Ediacaran-Cambrian, suggesting that the Proterozoic dynamo might have experienced a progressive decay when compositional convection was lacking and thermal convection was decreasing \cite{Driscoll2016a, Davies2021a}.

@@

However, previous paleointensity data from intrusive and extrusive rocks in the \textit{ca.} 1.1 Ga Midcontinent Rift \cite{Pesonen1983a, Kulakov2013a, Sprain2018a} record VDM values similar to that of today, which are higher than the values predicted by the long term decay curve proposed by ref. \citealp{Bono2019a} (Fig. \ref{fig:PINT_compilation}). Ref. \citealp{Pesonen1983a} obtained an average virtual dipole moment of 93 $\times 10^{21}$ Am$^2$ (ZAm$^2$) after applying a 13\% cooling rate correction (114 ZAm$^2$ before correction) using both extrusive and intrusive rocks from the Midcontinent Rift, including the \textit{ca.} 1108-1105 Ma Osler Volcanics, the \textit{ca.} 1100 Ma Maimainse Point volcanics, the \textit{ca.} 1107 Ma intrusive Logan Sills, Thunder Bay dikes, the Baraga and Marquette dikes, as well as lava flows of the \textit{ca.} 1096 Ma North Shore Volcanic Group (Fig. \ref{fig:Geologic_map}). Ref. \citealp{Kulakov2013a} obtained an average VDM of 59 $\pm$ 11 ZAm$^2$ using the \textit{ca.} 1085 Ma Lake Shore Traps volcanics and suggest that the present-day-like geomagnetic field strength likely existed through the late stage of Midcontinent Rift magmatism \cite{Vervoort2007a, Miller2013a}. However, non-ideal paleointensity behaviors at specimen level and overall low success rates have challenged the interpretation of many previous paleointensity results obtained by ref. \citealp{Pesonen1983a, Kulakov2013a}. One typical failure of paleointensity experiment is the double-slope behavior, as two distinct paleointensity estimates may be calculated depending on the interpreter's choice of slope: a higher paleointensity estimate from the low-temperature portion and a lower paleointensity estimate from the high-temperature portion. Ref \citealp{Pesonen1983a} used the low-temperature slope as the best representation of the past magnetic field strength (likely overestimating the field strength) whereas ref. \citealp{Kulakov2013a} used the high-temperature slope (likely underestimating the field strength). Results with such non-ideal behavior were rejected by ref. \citealp{Sprain2018a} who applied stricter paleointensity selection criteria on new paleointensity results and accepted only 23 out of 133 specimens (5 out of 21 lava flows) from the Osler Volcanics and the Mamainse Point volcanics. A calculated mean VDM value of 56 $\pm$ 21 ZAm$^2$ is similar to the average value of the past 300 million years \cite{Sprain2018a}.

Although it is typical of paleointensity experiments to have low success rates, bulk plagioclase-rich rocks \cite{Selkin2000a} and single plagioclase crystals \cite{Tarduno2005a} are found to be able to produce high quality paleointensity results. It has been suggested that plagioclase crystals can protect magnetic minerals that exsolved within them against post-formation weathering and thermochemical alteration during paleointensity experiments \cite{Tarduno2005a}. In this study, we present new paleointensity results from IZZ modified absolute paleointensity experiments \cite{Yu2004a} on a suite of bulk samples from \textit{ca.} 1092 Ma anorthosite xenoliths and their Beaver River diabase host rocks of the Beaver Bay Complex within the Midcontinent Rift. Paleointensity experiments on the anorthosite xenoliths have a very high success rate. Their consistent specimen- and site-level paleointensity results together with rock magnetic data reveal that the anorthosite xenoliths have low anisotropy of thermal remanent magnetization (TRM) and can acquire TRM linearly. Magnetic imaging shows the anorthosite specimens have dominant magnetic carriers interstitial to and within plagioclase crystals and they do not display strong preferred orientations. Our single-slope, high quality paleointensity data add to the reliable paleointensity records during the $\sim$25 Myr of Midcontinent Rift magmatic activity. Our new data from the Midcontinent Rift intrusive rocks support previous data from the volcanics that there was a strong late Mesoproterozoic geodynamo.

@@

The Mesoproterozoic North American Midcontinent Rift is an intracontinental rift that did not result in the break-up of Laurentia. Although the protracted magmatic activity in the rift lasted form \textit{ca.} 1109 Ma to \textit{ca.} 1084 Ma \cite{Swanson-Hysell2019a}, its activity during the main magmatic stage ($\sim$1098–1090 Ma; refs. \citealp{Vervoort2007a, Miller2013a}) was characterized by rapid and voluminous emplacement of intrusive rocks and extrusive volcanics. The $<$500 Kyr emplacement of the \textit{ca.} 1096 Ma Duluth Complex and the associated $\sim$8 km thick extrusive volcanics is one manifestation \cite{Swanson-Hysell2021a}. The emplacement of the \textit{ca.} 1092 Ma Beaver Bay Complex in northern Minnesota punctuates another such period of magmatism during the Midcontinent Rift main stage activity. The hypabyssal intrusion of the 1091.7 $\pm$ 0.2 Ma Beaver River diabase dikes and sills of the Beaver Bay Complex transported numerous anorthosite xenoliths that have short-axis diameters up to 180 meters via wide conduits \cite{Boerboom2004a, Boerboom2006b}. Geochronology, geochemical, and paleomagnetic data support the hypothesis that the diabase extruded to the surface and fed the coeval voluminous Greenstone Flow of the Portage Lake volcanics \cite{Doyle2016a, Zhang2021b}.

Thermal modeling results and paleomagnetic directional data show the anorthosite xenoliths acquired thermal remanent magnetizations during cooling within the Beaver River diabase \cite{Zhang2021b}. Step-wise thermal demagnetizations show the anorthosite xenoliths have minimal secondary remanence and single-component characteristic remanence magnetizations that often unblock sharply within temperature ranges between 500\textdegree C and 580\textdegree C. Geochronology data and rare earth element diffusion modeling further suggest that the anorthosite xenoliths are cumulate enclaves that formed at the time of Beaver Bay Complex magmatism. Petrological textures and geochemical attributes imply that the Beaver River anorthosite xenoliths formed in a lower crustal magmatic conditionm \cite{Zhang2021b, Miller1997a} distinct from that of layered mafic intrusions, such as the Duluth Complex which is $\sim$4 Myr older and generated massive anorthositic cumulates that led to the formation of the Anorthositic Series \cite{Swanson-Hysell2021a, Miller1990a}. The dominantly monomineralic anorthosite xenoliths of the Beaver River diabase have an average anorthite content of $\sim$70\% \cite{Morrison1983a, Doyle2016a}, higher than the average of $\sim$60\% of the Anorthositic Series rocks of the Duluth Complex \cite{Miller1990a}. The plagioclase in the Duluth Complex Anorthositic Series typically are euhedral with an interlocking texture showing igneous foliations (Fig. \ref{fig:Petro_QDM}A), but the more calcium-rich plagioclase of the Beaver River anorthosites often develop granoblastic texture characterized by equigranular crystals with poor alignments (Fig. \ref{fig:Petro_QDM}B). Abundant magnetite and ilmenite needles that are typically 10 $\mu$m long can be found to have preferred orientations within Duluth Complex plagioclase. Interstitial titanomagnetite and ilmenite symplectic intergrowths also coexist with pyroxene and relict olivine as a product of olivine oxidation in the Duluth Complex plagioclase \cite{Miller1990a}. In contrast, titanomagnetite needles and the symplectic magnetite-ilmenite intergrowths are absent but relatively sparse titano-magnetite within or interstitial to the plagioclase are typical of Beaver River anorthosites (Fig. \ref{fig:Petro_QDM}). This reflects that the Beaver River diabase magma had a different oxidation history than that of the Duluth Complex. The plagioclase and oxide mineral textures of the Beaver River anorthosites also distinguish them from plagioclase cumulates of other layered mafic complex where magnetic mineral fabrics associated with igneous foliation often occur \cite{Scofield1986a, Selkin2000a, Feinberg2006a}.

@@

\begin{figure*}
\centering
\noindent\includegraphics[width=11.4 cm]{Geologic_map.pdf}
\caption{\footnotesize{Simplified geologic map of Lake Superior region showing the distribution of rift and post-rift-related rocks including extrusive, intrusive, and late/post-rift sediments around Lake Superior. Red diamonds mark sites with paleointensity results that passed the selection criteria from this study. Paleomagnetic sites from ref. \citealp{Pesonen1983a} are categorized by lithology. All sites from refs. \citealp{Pesonen1983a, Kulakov2013a, Sprain2018a} are color-coded by their ages.}}
\label{fig:Geologic_map}
\end{figure*}

@@

\begin{figure*}
\centering
\noindent\includegraphics[width=11.4 cm]{Petro_QDM.png}
\caption{\footnotesize{Thin section petrographic and magnetic images of anorthosites from Duluth Complex Anorthositic Series (left column) and Beaver River anorthosite xenolith from which paleomagnetic site AX16 and geochronology sample MS99033 were collected (right column; ref. \citealp{Zhang2021b}). Petrographic image (A) shows Duluth Complex anorthosite have euhedral, interlocking plagioclase crystals showing igneous foliation. (B) Beaver River anorthosite have plagioclase with a granoblastic texture. (C) Duluth Complex plagioclase can contain abundant Fe-Ti oxide needles that have preferred orientations that are often parallel with the (010) planes of plagioclase. In contrast, plagioclase of sample MS99033 do not enclose large oxide needles. Magnetic field maps acquired by the quantum diamond microscope (QDM) at UC Berkeley Paleomagnetism Lab show that the oxide needles within Duluth Complex plagioclase are strongly magnetic as evidenced through the complex overlapping magnetic fields exerted above the oxides in (E). In comparison, relatively weak magnetic sources within and interstitial to plagioclase crystals exist in the Beaver River anorthosite. The dipole-like features in (F) do not show strong preferred orientations. Magnetic field maps of (E) and (F) share the same spatial and magnetic scale. The unit of the axes are in $\mu$m.}}
\label{fig:Petro_QDM}
\end{figure*}

>>>>

% Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
\author[a,1]{Yiming Zhang}
\author[a]{Nicholas L. Swanson-Hysell} 
\author[a,b]{Margaret S. Avery}
\author[c]{Roger R. Fu}

@@

% Please add a significance statement to explain the relevance of your work
\significancestatement{Acquiring high-fidelity observational records of ancient magnetic field intensity from remanent magnetization of rocks is crucial for constraining the evolution history of Earth's core. However, robust estimates of ancient field strengths are often difficult to recover due to alterations or non-ideal rock magnetic behaviors. In this study, we use plagioclase cumulates that formed in the deeper crust and were brought to the near surface by magma where they cooled and acquired thermal remanent magnetizations. These plagioclase from the North American Midcontinent Rift have experienced minimal post-formation alterations and record a present-day-like virtual dipole moment 1.1 billion years ago. The results indicate the existence of strong power sources for Earth's dynamo during the late Mesoproterozoic. }

% Please include corresponding author, author contribution and author declaration information
\authorcontributions{Author contributions: Y.Z. and N.L.S.-H. designed research; Y.Z., N.L.S.-H. and M.S.A conducted fieldwork and sampling; Y.Z. and N.L.S.-H. conducted and analyzed rock magnetic experiments; Y.Z. performed petrography and electron microscopy; R.R.F. contributed to magnetic imaging data analyses; Y.Z. and N.L.S.-H. wrote the paper.}

@@

\begin{abstract}
Obtaining estimates of Earth's magnetic field strength in deep time is complicated by non-ideal rock magnetic behavior in many igneous rocks. In this study, we target \textit{ca.} 1092 Ma anorthosite xenoliths that cooled and acquired their magnetization within Beaver River diabase intrusions of the North American Midcontinent Rift. In contrast to the diabase which has non-ideal paleointensity behaviors and fails to provide reliable paleointensity estimates, the anorthosite xenoliths are unusually high fidelity recorders yielding high-quality, single-slope paleointensity results that are consistent at specimen and site levels. An average value of $\sim$89 ZAm$^2$ for the virtual dipole moment from the anorthosites is higher than that of the dipole component of Earth's magnetic field today. Such a strong late Mesoproterozoic dipole recorded by the anorthosites requires the existence of a vigorous geodynamo at the time. Together with previous paleointensity data from other Midcontinent Rift rocks, our new results indicate that a dynamo with strong power sources persisted through $\sim$25 Myr at \textit{ca.} 1.1 Ga. These observations are inconsistent with there being a progressive monotonic decay of Earth's dynamo strength through the Proterozoic Eon.
\end{abstract}

@@

% If your first paragraph (i.e. with the \dropcap) contains a list environment (quote, quotation, theorem, definition, enumerate, itemize...), the line after the list may have some extra indentation. If this is the case, add \parshape=0 to the end of the list environment.
\dropcap{E}arth's magnetic field is the result of convective flow of liquid iron-alloy in Earth's outer core. At present day, the geodynamo is collectively powered by heat flow across the core-mantle boundary (CMB), and from the crystallization of the solid inner core from the liquid outer core which provides latent heat and compositional buoyancy. However, while paleomagnetic studies have found that a dynamo field has existed since at least 3.4 billion years ago \cite{Selkin2007a, Tarduno2014a, Brenner2020a}, Earth's inner core likely crystallized more recently. Estimates of the timing of the initial crystallization of Earth's inner core are interconnected with estimates for the core's thermal conductivity. Higher conductivity values imply faster cooling rates, which in turn imply that the threshold for the freezing of the inner core happened more recently \cite{Davies2015a}. While some estimates of these values are consistent with an inner core age $>$3 Ga \cite{Gubbins2004a, Konopkova2016a}, other estimates have implied higher thermal conductivity values and an age for the inner core that is less than 1.3 Ga \cite{Pozzo2012a, Koker2012a, Gomi2013a, Zhang2020b}, with some suggesting even younger ages ($<$ 700 Ma; ref. \citealp{Labrosse2015a, Ohta2016a}). Given that estimates for the core's thermal conductivity continue to be debated, it is crucial to use observational records as an independent constraint on the thermal evolution of Earth's core and mantle.

@@

Paleomagnetic records from ancient rocks are one of the few types of observational data that have the potential to record the thermal evolution of Earth’s core. Evidence for a persistent magnetic field through the Proterozoic, for example, necessitates the existence of plate tectonics that sustained core-mantle boundary heat flow. However, strikingly low estimates of geomagnetic field strengths have been obtained in the Ediacaran Period (ref. \citealp{Bono2019a}; \textit{ca.} 565 Ma), potentially indicating an unusual period of geodynamo regime at the time.

% Those paleointensity data has been interpreted to be consistent with thermal evolution models that use high core thermal conductivity values, where the nucleation of the inner core did not begin to provide compositional convection power for driving the geodynamo until after \textit{ca.} 565 Ma, suggesting that the Proterozoic dynamo might have experienced a monotonic decay when compositional convection was lacking and thermal convection was progressively decreasing \cite{Driscoll2016a, Davies2021a}.

Determinations of ancient magnetic field strengths rely on igneous rocks that acquire thermal remanent magnetizations as they cool in Earth's magnetic field. These magnetizations need to be unmodified by subsequent heating or chemical alterations in order to be able to faithfully record ancient geomagnetic field record through time. Intraplate magmatic events are therefore an important target as they can be well-preserved within continental interiors at distance from mountain-building events along continental margins. However, magmatic events associated with large igneous provinces are typically of geologically short duration with the bulk of magmatic products emplaced within $<$1 Myr. The Midcontinent Rift is an exception as magmatism lasted $\sim$25 Myr from 1109 Ma to 1084 Ma. Additionally, extension ceased in the Midcontinent Rift prior to lithosphere separation, preserving volcanics, intrusives, and sediments of the rift within the continental interior. As a result, rocks of the rift have simple paleomagnetic behavior and form a central record of Mesoproterozoic paleogeography \cite{Swanson-Hysell2021c}. This simple magnetization and the long duration of magnetization holds the potential for a well-preserved extended window into the intensity of Earth's magnetic field from 1109 to 1084 Ma.

Despite the exquisite preservation of the rocks, non-ideal paleointensity behaviors and overall low success rates have often led to high failure rates and challenged the interpretation of many previous paleointensity results from the Midcontinent Rift \cite{Pesonen1983a, Kulakov2013a}. One typical failure of paleointensity experiment is the double-slope behavior, as two distinct paleointensity estimates may be calculated depending on the interpreter's choice of slope: a higher paleointensity estimate from the low-temperature portion and a lower paleointensity estimate from the high-temperature portion. Ref \citealp{Pesonen1983a} used the low-temperature slope as the best representation of the past magnetic field strength (likely overestimating the field strength) whereas ref. \citealp{Kulakov2013a} used the high-temperature slope (likely underestimating the field strength). Results with such non-ideal behavior were rejected by ref. \citealp{Sprain2018a} who applied stricter paleointensity selection criteria.

In this study, we target a unique rock type - high-purity anorthosite xenoliths. These anorthosite xenoliths were brought to the near surface with magma that formed sills of the Beaver River diabase. They cooled and acquired their magnetizations in conjunction with the host diabase. Paleointensity experiments on the anorthosite xenoliths have a very high success rate, yielding consistent specimen- and site-level paleointensity results. Rock magnetic data reveal that the anorthosite xenoliths have low anisotropy of thermal remanent magnetization (TRM) and can acquire TRM linearly. Magnetic imaging shows the anorthosite specimens have dominant magnetic carriers interstitial to and within plagioclase crystals and they do not display strong preferred orientations. Our single-slope, high quality paleointensity data add to the reliable paleointensity records during the $\sim$25 Myr of Midcontinent Rift magmatic activity. Our new data from the Midcontinent Rift intrusive rocks support previous data from the volcanics that there was a strong late Mesoproterozoic geodynamo.

% Interpretations of selected records of the current Precambrian absolute paleointensity database (compiled in the PINT database; \url{http://earth.liv.ac.uk/pint/}) suggest that an appreciable geomagnetic field existed in the Archean but reached its weakest in the Ediacaran when very low virtual dipole moment (VDM) estimates have been obtained \cite{Bono2019a, Shcherbakova2019a, Thallner2021b, Lloyd2021a}, followed by a recovery near the Ediacaran-Cambrian boundary \cite{Thallner2021a}. Ref. \citealp{Bono2019a} adopted a billion-year-scale view of such observations to suggest that these data are consistent with thermal evolution models that use high core thermal conductivity values, where the nucleation of the inner core did not begin to provide compositional convection power for driving the geodynamo until the Ediacaran-Cambrian, suggesting that the Proterozoic dynamo might have experienced a monotonic decay when compositional convection was lacking and thermal convection was progressively decreasing \cite{Driscoll2016a, Davies2021a}.

@@

% However, previous paleointensity data from intrusive and extrusive rocks in the \textit{ca.} 1.1 Ga Midcontinent Rift \cite{Pesonen1983a, Kulakov2013a, Sprain2018a} record VDM values similar to that of today, which are higher than the values predicted by the long term decay curve proposed by ref. \citealp{Bono2019a} (Fig. \ref{fig:PINT_compilation}). Ref. \citealp{Pesonen1983a} obtained an average virtual dipole moment of 93 $\times 10^{21}$ Am$^2$ (ZAm$^2$) after applying a 13\% cooling rate correction (114 ZAm$^2$ before correction) using both extrusive and intrusive rocks from the Midcontinent Rift, including the \textit{ca.} 1108-1105 Ma Osler Volcanics, the \textit{ca.} 1100 Ma Maimainse Point volcanics, the \textit{ca.} 1107 Ma intrusive Logan Sills, Thunder Bay dikes, the Baraga and Marquette dikes, as well as lava flows of the \textit{ca.} 1096 Ma North Shore Volcanic Group (Fig. \ref{fig:Geologic_map}). Ref. \citealp{Kulakov2013a} obtained an average VDM of 59 $\pm$ 11 ZAm$^2$ using the \textit{ca.} 1085 Ma Lake Shore Traps volcanics and suggest that the present-day-like geomagnetic field strength likely existed through the late stage of Midcontinent Rift magmatism \cite{Vervoort2007a, Miller2013a}. However, non-ideal paleointensity behaviors at specimen level and overall low success rates have challenged the interpretation of many previous paleointensity results obtained by ref. \citealp{Pesonen1983a, Kulakov2013a}. One typical failure of paleointensity experiment is the double-slope behavior, as two distinct paleointensity estimates may be calculated depending on the interpreter's choice of slope: a higher paleointensity estimate from the low-temperature portion and a lower paleointensity estimate from the high-temperature portion. Ref \citealp{Pesonen1983a} used the low-temperature slope as the best representation of the past magnetic field strength (likely overestimating the field strength) whereas ref. \citealp{Kulakov2013a} used the high-temperature slope (likely underestimating the field strength). Results with such non-ideal behavior were rejected by ref. \citealp{Sprain2018a} who applied stricter paleointensity selection criteria on new paleointensity results and accepted only 23 out of 133 specimens (5 out of 21 lava flows) from the Osler Volcanics and the Mamainse Point volcanics. A calculated mean VDM value of 56 $\pm$ 21 ZAm$^2$ is similar to the average value of the past 300 million years \cite{Sprain2018a}.

% Although it is typical of paleointensity experiments to have low success rates, bulk plagioclase-rich rocks \cite{Selkin2000a} and single plagioclase crystals \cite{Tarduno2005a} are found to be able to produce high quality paleointensity results. It has been suggested that plagioclase crystals can protect magnetic minerals that exsolved within them against post-formation weathering and thermochemical alteration during paleointensity experiments \cite{Tarduno2005a}. In this study, we present new paleointensity results from IZZ modified absolute paleointensity experiments \cite{Yu2004a} on a suite of bulk samples from \textit{ca.} 1092 Ma anorthosite xenoliths and their Beaver River diabase host rocks of the Beaver Bay Complex within the Midcontinent Rift. Paleointensity experiments on the anorthosite xenoliths have a very high success rate. Their consistent specimen- and site-level paleointensity results together with rock magnetic data reveal that the anorthosite xenoliths have low anisotropy of thermal remanent magnetization (TRM) and can acquire TRM linearly. Magnetic imaging shows the anorthosite specimens have dominant magnetic carriers interstitial to and within plagioclase crystals and they do not display strong preferred orientations. Our single-slope, high quality paleointensity data add to the reliable paleointensity records during the $\sim$25 Myr of Midcontinent Rift magmatic activity. Our new data from the Midcontinent Rift intrusive rocks support previous data from the volcanics that there was a strong late Mesoproterozoic geodynamo.

@@

While magmatic activity within the Midcontinent Rift was protracted, there were volcanism and intrusions characterized by particularly rapid and voluminous emplacement. The emplacement of the \textit{ca.} 1092 Ma Beaver Bay Complex in northern Minnesota punctuates one such period of magmatism during the Midcontinent Rift main stage activity. The magma that formed the 1091.7 $\pm$ 0.2 Ma Beaver River diabase dikes and sills of the Beaver Bay Complex transported numerous anorthosite xenoliths that have short-axis diameters up to 180 meters via wide conduits \cite{Boerboom2004a, Boerboom2006b}. The anorthosite xenoliths are plagioclase cumulates that formed comagmatically with the host diabase in the lower crust - an interpretation confirmed by U-Pb zircon geochronology \cite{Zhang2021b}.

% The Mesoproterozoic North American Midcontinent Rift is an intracontinental rift that did not result in the break-up of Laurentia. Although the protracted magmatic activity in the rift lasted form \textit{ca.} 1109 Ma to \textit{ca.} 1084 Ma \cite{Swanson-Hysell2019a}, its activity during the main magmatic stage ($\sim$1098–1090 Ma; refs. \citealp{Vervoort2007a, Miller2013a}) was characterized by rapid and voluminous emplacement of intrusive rocks and extrusive volcanics. The $<$500 Kyr emplacement of the \textit{ca.} 1096 Ma Duluth Complex and the associated $\sim$8 km thick extrusive volcanics is one manifestation \cite{Swanson-Hysell2021a}. The emplacement of the \textit{ca.} 1092 Ma Beaver Bay Complex in northern Minnesota punctuates another such period of magmatism during the Midcontinent Rift main stage activity. The hypabyssal intrusion of the 1091.7 $\pm$ 0.2 Ma Beaver River diabase dikes and sills of the Beaver Bay Complex transported numerous anorthosite xenoliths that have short-axis diameters up to 180 meters via wide conduits \cite{Boerboom2004a, Boerboom2006b}. Geochronology, geochemical, and paleomagnetic data support the hypothesis that the diabase extruded to the surface and fed the coeval voluminous Greenstone Flow of the Portage Lake volcanics \cite{Doyle2016a, Zhang2021b}.

Thermal modeling results and paleomagnetic directional data show the anorthosite xenoliths to have acquired thermal remanent magnetizations while cooing with the Beaver River diabase \cite{Zhang2021b}. Step-wise thermal demagnetizations show the anorthosite xenoliths have dominantly single-component magnetization that often unblock sharply within temperature ranges between 500\textdegree C and 580\textdegree C.

\section*{Petrography and magnetic imaging of anorthosite xenoliths}

Petrographic imaging shows the dominantly monomineralic anorthosite xenoliths of the Beaver River diabase often develop granoblastic texture characterized by equigranular crystals with poor development of petrofabrics (Fig. \ref{fig:Petro_QDM}A). In addition, the plagioclase crystals of the anorthosite xenoliths are often free of large inclusions. Magnetic imaging using the quantum diamond microscope (QDM) shows that the dominant magnetic remanence carriers often exist within and sometimes interstitial to plagioclase crystals (Fig. \ref{fig:Petro_QDM}E).

These features of the Beaver River anorthosite xenoliths are distinct from other anorthosites such as the anorthosite xenoliths within the Duluth Complex Anorthositic Series rocks. The plagioclase of the Duluth Complex anorthosites typically develop interlocking textures that display strong igneous foliations (Fig. \ref{fig:Petro_QDM}B) and associated magnetic mineral fabrics characterized by having aligned titanomagnetite needles within plagioclase crystals (Fig. \ref{fig:Petro_QDM}D). Magnetic imaging confirms that the titanomagnetite needles within the plagioclase crystals are often strongly magnetic with magnetic fabrics oriented along the long axes of the needles (Fig. \ref{fig:Petro_QDM}F). Titanomagnetite and ilmenite symplectic intergrowths also coexist with pyroxene and relict olivine as a product of olivine oxidation in the Duluth Complex plagioclase \cite{Miller1990a}.

The lack of symplecites and exsolved titanomagnetite needles within the Beaver River anorthosite xenoliths reflects that the Beaver River diabase magma had a different oxidation history than that of the Duluth Complex. The plagioclase and oxide mineral textures of the Beaver River anorthosites also distinguish them from plagioclase cumulates of other layered mafic complex where magnetic mineral fabrics associated with igneous foliation often occur \cite{Scofield1986a, Selkin2000a, Feinberg2006a}.

% The dominantly monomineralic anorthosite xenoliths of the Beaver River diabase have an average anorthite content of $\sim$70\% \cite{Morrison1983a, Doyle2016a}, higher than the average of $\sim$60\% of the Anorthositic Series rocks of the Duluth Complex \cite{Miller1990a}. The plagioclase in the Duluth Complex Anorthositic Series typically are euhedral with an interlocking texture showing igneous foliations (Fig. \ref{fig:Petro_QDM}A), but the more calcium-rich plagioclase of the Beaver River anorthosites often develop granoblastic texture characterized by equigranular crystals with poor alignments (Fig. \ref{fig:Petro_QDM}B). Abundant magnetite and ilmenite needles that are typically 10 $\mu$m long can be found to have preferred orientations within Duluth Complex plagioclase. Interstitial titanomagnetite and ilmenite symplectic intergrowths also coexist with pyroxene and relict olivine as a product of olivine oxidation in the Duluth Complex plagioclase \cite{Miller1990a}. In contrast, titanomagnetite needles and the symplectic magnetite-ilmenite intergrowths are absent but relatively sparse titano-magnetite within or interstitial to the plagioclase are typical of Beaver River anorthosites (Fig. \ref{fig:Petro_QDM}). This reflects that the Beaver River diabase magma had a different oxidation history than that of the Duluth Complex. The plagioclase and oxide mineral textures of the Beaver River anorthosites also distinguish them from plagioclase cumulates of other layered mafic complex where magnetic mineral fabrics associated with igneous foliation often occur \cite{Scofield1986a, Selkin2000a, Feinberg2006a}.

@@

\begin{figure*}
\centering
\noindent\includegraphics[width=11.4 cm]{Geologic_map.pdf}
\caption{\footnotesize{Simplified geologic map of Lake Superior region showing the distribution of rift and post-rift-related rocks including extrusive, intrusive, and late/post-rift sediments around Lake Superior. Red diamonds mark sites with paleointensity results that passed the selection criteria from this study. Paleomagnetic sites from ref. \citealp{Pesonen1983a} are categorized by lithology. All sites from ref. \citealp{Pesonen1983a, Kulakov2013a, Sprain2018a} are color-coded by their ages.}}
\label{fig:Geologic_map}
\end{figure*}

@@

\begin{figure*}
\centering
\noindent\includegraphics[width=11.4 cm]{Petro_QDM.png}
\caption{\footnotesize{Thin section petrographic and magnetic images of anorthosites from the Beaver River anorthosite xenolith from which paleomagnetic site AX16 and geochronology sample MS99033 were collected (left column; ref. \citealp{Zhang2021b}) and and distinct anorthosite from the Duluth Complex Anorthositic Series (right column). Cross-polarized petrographic images (A) Beaver River anorthosite have plagioclase with a granoblastic texture; (B) Plagioclase of Duluth Complex anorthosite have euhedral, interlocking crystals showing igneous foliation; (C) zoomed-in image of plagioclase from MS99033 showing the crystals are largely free of large opaque inclusions; (D) Duluth Complex plagioclase can contain abundant Fe-Ti oxide needles that have preferred orientations that are often parallel with the (010) planes of plagioclase. Magnetic field maps acquired by the quantum diamond microscope (QDM) show that the relatively weak magnetic sources within and interstitial to plagioclase crystals exist in the Beaver River anorthosite in (E) while large oxide needles within Duluth Complex plagioclase are strongly magnetic as evidenced through the complex overlapping magnetic fields above the oxides in (F). Magnetic field maps of (E) and (F) share the same spatial and magnetic scale. The unit of the axes are in $\mu$m.}}
\label{fig:Petro_QDM}
\end{figure*}
